
Analysis and visualization of expert assessments of orangutan population trends under different scenarios. For Meijaard et al. manuscript "Expert guided analysis for restoring the red ape in a whole or half earth context"

2022-04-11
Andy Marshall

# Load packages, versions

R version 4.1.3 (2022-03-10) 

tidyverse 1.3.1 ──
✓ ggplot2 3.3.5     ✓ purrr   0.3.4
✓ tibble  3.1.6     ✓ dplyr   1.0.8
✓ tidyr   1.2.0     ✓ stringr 1.4.0
✓ readr   2.1.2     ✓ forcats 0.5.1
ggpubr 0.4.0

```{r}

library(tidyverse)
library(ggpubr)

```

# Pull & wrangle data

```{r}

d <- read_csv("data/data-2022-03-23.csv")
d <- d %>%
     pivot_longer(cols = -Names, names_to = 'Expert') %>%
     pivot_wider(names_from = Names, values_from = value)

```

# Analysis

##Variation among experts

```{r}

# Absolute numbers quite variable:
mean(d$pop_current)
range(d$pop_current)
mean(d$offtake_current, na.rm = TRUE)
range(d$offtake_current, na.rm = TRUE)

# Relative effects broadly consistent, e.g.,
range(d$pop_2032_WE - d$pop_2032_current)
range(d$pop_2032_HE - d$pop_2032_current) 
range(d$offtake_WE - d$offtake_current, na.rm = TRUE) 
range(d$offtake_HE - d$offtake_current, na.rm = TRUE)

```

## Current conditions

```{r}

d$per_left_nochange <- 100 * d$pop_2032_current/d$pop_current
median(d$per_left_nochange)
quantile(d$per_left_nochange)

d$offtake_rate_current <- 100 * (d$offtake_current / d$pop_current)
median(d$offtake_rate_current, na.rm = TRUE)
quantile(d$offtake_rate_current, na.rm = TRUE)

```

## Half Earth

```{r}

d$per_left_HE       <- 100 * d$pop_2032_HE/d$pop_current
median(d$per_left_HE)
range(d$per_left_HE)

# all but one expert predicted decline less than predicted under current conditions:
d$per_left_HE - d$per_left_nochange

d$offtake_rate_HE      <- 100 * (d$offtake_HE / d$pop_current)
median(d$offtake_rate_HE)
range(d$offtake_rate_HE)

```

## Whole Earth

```{r}

d$per_left_WE       <- 100 * d$pop_2032_WE/d$pop_current
median(d$per_left_WE)
range(d$per_left_WE)

d$offtake_rate_WE      <- 100 * (d$offtake_WE / d$pop_current)
median(d$offtake_rate_WE)
range(d$offtake_rate_WE)


```

## Aspriational scenario

```{r}

d$per_2122          <- 100 * d$pop_2122/d$pop_current
median(d$per_2122)
range(d$per_2122)

```



# Figures

## Figure 02

```{r}

col.NC <- "cornflowerblue"
col.HE <- "darkorange"
col.WE <- "darkorange4"
col.DR <- "darkgreen"

d$per_left_HE       <- 100 * d$pop_2032_HE/d$pop_current
d$per_left_nochange <- 100 * d$pop_2032_current/d$pop_current
d$per_left_WE       <- 100 * d$pop_2032_WE/d$pop_current
d$per_2122          <- 100 * d$pop_2122/d$pop_current

p1 <-ggplot(d, aes(x = per_left_WE)) +
     geom_histogram(aes(y = ..density..), bins = 80,
                    col = "gray60", fill = "gray80") +
     geom_density(alpha = 0.6, fill = col.WE) +
     geom_point(aes(x=median(per_left_WE), y=0), 
                pch = 17, colour = "gray40", size = 3) +
     geom_point(aes(x=median(per_left_WE), y=0), 
                pch = 17, colour = "white", size = 1.5) +
     geom_segment(aes(x = 100, y = -0.001,
                      xend = 100, yend = 0.04),
                      color = "gray40", lty = 2, lwd = 0.65) +
     xlab("") +
     ylab("\n") + 
     scale_x_continuous(limits = c(-10, 520),
                        breaks = c(0, 100, 200, 300, 400, 500)) +
     scale_y_continuous(limits =c(-0.0025, 0.04),
                        breaks = c(0, 0.01, 0.02, 0.03, 0.04)) +
     theme_minimal() +
     theme(axis.text.x = element_blank()) +
     annotate(geom = "text", x = 410, y = 0.0375, cex = 4.5,
              hjust = 0, label = "Whole Earth scenario", col = col.WE) +
     annotate(geom = "text", x = 410, y = 0.0325, cex = 4.5,
              hjust = 0,label = "(2032)", col = col.WE) +
     annotate(geom = "text", x = median(d$per_left_WE), y = -0.0025, 
              cex = 3.5, label = "median", col = "gray40") +
     annotate(geom = "text", x = -10, y = 0.0375, 
              cex = 7, label = "a)", col = "black")
    
p2 <- ggplot(d, aes(x = per_left_nochange)) +
     geom_histogram(aes(y = ..density..), bins = 80,
                    col = "gray60", fill = "gray80") +
     geom_density(alpha = 0.6, fill = col.NC) +
     geom_segment(aes(x = 100, y = -0.001,
                      xend = 100, yend = 0.04),
                      color = "gray40", lty = 2, lwd = 0.65) +    
     geom_point(aes(x = median(per_left_nochange), y = 0), 
                pch = 17, colour="gray40", size = 3) +
     geom_point(aes(x = median(per_left_nochange), y = 0), 
                pch = 17, colour="white", size = 1.5) +
     xlab("") +
     ylab("\n") + 
     scale_x_continuous(limits = c(-10, 520),
                        breaks = c(0, 100, 200, 300, 400, 500)) +
     scale_y_continuous(limits =c(-0.001, 0.04),
                        breaks = c(0, 0.01, 0.02, 0.03, 0.04)) +
     theme_minimal() +
     theme(axis.text.x = element_blank()) +
     annotate(geom = "text", x = 410, y = 0.0375, cex = 4.5,
              hjust = 0, label = "Current conditions", col = col.NC) +
     annotate(geom = "text", x = 410, y = 0.0325, cex = 4.5,
              hjust = 0, label = "(2032)", col = col.NC)+
     annotate(geom = "text", x = -10, y = 0.0375, 
             cex = 7, label = "b)", col = "black")

p3 <- ggplot(d, aes(x = per_left_HE)) +
     geom_histogram(aes(y = ..density..), bins = 80,
                    col = "gray60", fill = "gray80") +
     geom_density(alpha = 0.6, fill = col.HE) +
     geom_segment(aes(x = 100, y = -0.001,
                      xend = 100, yend = 0.04),
                      color = "gray40", lty = 2, lwd = 0.65) +
     geom_point(aes(x = median(per_left_HE), y = 0), 
                pch = 17, colour = "gray40", size = 3) +
     geom_point(aes(x = median(per_left_HE), y = 0), 
                pch = 17, colour="white", size = 1.5) +    
     theme_minimal() +
     theme(axis.text.x = element_blank()) +
     xlab("") +
     ylab("\n") + 
     scale_x_continuous(limits = c(-10, 520),
                        breaks = c(0, 100, 200, 300, 400, 500)) +
     scale_y_continuous(limits =c(-0.001, 0.04),
                        breaks = c(0, 0.01, 0.02, 0.03, 0.04)) +
     annotate(geom = "text", x = 410, y = 0.0375, cex = 4.5,
              hjust = 0, label = "Half Earth scenario", col = col.HE) +
     annotate(geom = "text", x = 410, y = 0.0325, cex = 4.5,
              hjust = 0, label = "(2032)", col = col.HE)+
     annotate(geom = "text", x = -10, y = 0.0375, 
              cex = 7, label = "c)", col = "black")
  
p4 <- ggplot(d, aes(x = per_2122)) +
     geom_histogram(aes(y = ..density..), bins = 80,
                    col = "gray60", fill = "gray80") +
     geom_density(alpha = 0.6, fill = col.DR) +
     geom_segment(aes(x = 100, y = -0.001,
                      xend = 100, yend = 0.04),
                      color = "gray40", lty = 2, lwd = 0.65) +
     geom_point(aes(x = median(per_2122), y = 0), 
                pch = 17, colour = "gray40", size = 3) +
     geom_point(aes(x=median(per_2122), y=0), 
                pch = 17, colour = "white", size = 1.5) + 
     theme_minimal() +
     xlim(-100, 400)+
     xlab("\nEstimated % of current population") + 
     ylab("\n") + 
     scale_x_continuous(limits = c(-10, 520),
                        breaks = c(0, 100, 200, 300, 400, 500)) +
     scale_y_continuous(limits =c(-0.001, 0.04),
                        breaks = c(0, 0.01, 0.02, 0.03, 0.04)) +
     annotate(geom = "text", x = 410, y = 0.0375, cex = 4.5,
              hjust = 0, label = "Zero-offtake,", col = col.DR) +
     annotate(geom = "text", x = 410, y = 0.0325, cex = 4.5,
              hjust = 0, label = "zero-forest loss scenario", 
              col = col.DR)  +
     annotate(geom = "text", x = 410, y = 0.0275, cex = 4.5,
              hjust = 0, label = "(2122)", col = col.DR)+
     annotate(geom = "text", x = -10, y = 0.0375, 
              cex = 7, label = "d)", col = "black")

pdf("output/fig.02.pdf", height = 12, width = 10)

library(ggpubr)
figure <- ggarrange(p1, p2, p3, p4, 
                    ncol = 1, nrow = 4)
annotate_figure(figure,left = text_grob("Density",
                                 rot = 90))
dev.off()

```

## Figure 03

```{r}

col.NC <- "cornflowerblue"
col.HE <- "darkorange"
col.WE <- "darkorange4"
col.DR <- "darkgreen"

d$offtake_rate_current <- 100 * (d$offtake_current / d$pop_current)
d$offtake_rate_HE      <- 100 * (d$offtake_HE / d$pop_current)
d$offtake_rate_WE      <- 100 * (d$offtake_WE / d$pop_current)

scenarios <- cbind(d$offtake_rate_HE, 
                   d$offtake_rate_current, 
                   d$offtake_rate_WE)

my.cols         <- c(col.HE, col.NC, col.WE)
my.axis.cols    <- c("white","white", "black")
my.median.cols  <- c("gray40","white", "white")

my.xlims <- c(-2, max(scenarios, na.rm = TRUE)+1)
my.ylims <- c(-0.05, 0.65)

my.labels <- c("Half Earth scenario",
               "Current conditions",
               "Whole Earth scenario")
my.years1 <- c("(2032)", "(2032)", "(2032)")
my.panels <- c("a)", "b)", "c)")

my.plots <- list()

for (i in 1:3) {
   p <- eval(substitute(  
     ggplot(d, aes(x = scenarios[, i])) +
     geom_histogram(aes(y = ..density..), 
                    col = "gray60", 
                    fill = "gray80", 
                    bins = 50) +
     geom_density(alpha = 0.6, 
                  fill = my.cols[i]) +
     geom_point(aes(x = median(scenarios[, i], na.rm = TRUE), y = 0),
                pch = 17, colour="gray40", size = 3) +
     geom_point(aes(x = median(scenarios[, i], na.rm = TRUE), y = 0),
                pch = 17, colour="white", size = 1.5) +  
     annotate(geom = "text", 
              x = median(scenarios[, i], na.rm = TRUE),
              y = -0.05, position = 1,
              cex = 3, label = "median", col = my.median.cols[i]) +
     xlab("") + 
     ylab("\n") + 
     xlim(my.xlims[1], my.xlims[2]) +
     ylim(my.ylims[1], my.ylims[2]) +
     theme_minimal() +
     theme(axis.text.x = element_text(colour = my.axis.cols[i]),
           axis.title.x = element_text(colour = my.axis.cols[i])) +    
     annotate(geom = "text", 
              x = my.xlims[2],  y = 0.65,
              cex = 4.5,
              hjust = 1, vjust = 1,
              label = my.labels[i], col = my.cols[i]) + 
     annotate(geom = "text", 
              x = my.xlims[2], y = 0.64,
              cex = 4.5,
              hjust = 1, vjust = 2.5,
              label = my.years1[i], col = my.cols[i]) +
     annotate(geom = "text", x = -1.95, y = 0.58, 
              cex = 7, label = my.panels[i], col = "black")

     , list(i = i)))
     my.plots[[i]] <- p 
}  

pdf("output/fig.03.pdf", height = 10, width = 10)

figure <- ggarrange(my.plots[[1]], my.plots[[2]], my.plots[[3]],
                    ncol = 1, nrow = 3)
annotate_figure(figure, left = text_grob("Density",
                                 rot = 90),
                        bottom = text_grob("Estimated offtake rate (%)"))
dev.off()

```


## Figure S02

```{r}

d$offtake_rate_current <- 100 * (d$offtake_current / d$pop_current)
d$offtake_rate_HE      <- 100 * (d$offtake_HE / d$pop_current)
d$offtake_rate_WE      <- 100 * (d$offtake_WE / d$pop_current)

d2 <- as.data.frame(cbind(d$Expert, d$offtake_rate_HE, 
                          d$offtake_rate_current, d$offtake_rate_WE))

colnames(d2) <- c("Expert", "offtake_rate_HE", 
                  "offtake_rate_current", "offtake_rate_WE")

scenario <- c("offtake_rate_HE", "offtake_rate_current", "offtake_rate_WE")
num <- 1:3

crib <- as.data.frame(cbind(scenario, num))

d3 <- d2 %>% 
     gather(2:4, key = "scenario", value = "estimate") %>% 
     left_join(crib)

d3$jit_num <- jitter(as.numeric(d3$num), amount = 0.035)
d3$estimate <- as.numeric(d3$estimate)

pdf("output/fig.s02.pdf", height = 6, width = 5)

ggplot(d3, aes(x = jit_num, y = estimate)) +
     geom_point(aes(color = factor(Expert)), size = 2.5) +
     geom_line(aes(group = factor(Expert),
                   color = factor(Expert)),
                   alpha = 0.5) +
     scale_color_viridis_d() + 
     xlab("") + 
     ylab("Estimated offtake rate (%)") + 
     theme_minimal() +
     theme(plot.margin = margin(1, 1, 1, 0.5, "cm")) +
     theme(legend.position = "none") +  
     theme(panel.grid.major.y = element_line(size = 0.5,
                                             color = "gray80")) +
     scale_x_continuous(breaks = c(1,2,3),
                        limits = c (0.75, 3.1),
                        labels = c("Half Earth", 
                                   "Current", 
                                   "Whole Earth")) +
     theme(axis.text = element_text(size = 11)) +
     geom_point(aes(x = 1, y = median(d3$estimate[d3$num == "1"], na.rm = TRUE)),
                pch = 17, colour= "gray40", size = 4) +
     geom_point(aes(x = 1, y = median(d3$estimate[d3$num == "1"], na.rm = TRUE)),
                pch = 17, colour="white", size = 2) +
     geom_point(aes(x = 2, y = median(d3$estimate[d3$num == "2"], na.rm = TRUE)),
                pch = 17, colour= "gray40", size = 4) +
     geom_point(aes(x = 2, y = median(d3$estimate[d3$num == "2"], na.rm = TRUE)),
                pch = 17, colour="white", size = 2) +
     geom_point(aes(x = 3, y = median(d3$estimate[d3$num == "3"], na.rm = TRUE)),
                pch = 17, colour= "gray40", size = 4) +
     geom_point(aes(x = 3, y = median(d3$estimate[d3$num == "3"], na.rm = TRUE)),
                pch = 17, colour="white", size = 2) +
     annotate(geom = "text", x = 0.78,
               y = median(d3$estimate[d3$num == "1"], na.rm = TRUE), 
               position = 2, cex = 3, label = "median", col = "gray40")
dev.off()

```
